---
title: "Learn_rCharts"
output: html_document
---

About rCharts
- a way to create interactive javascript visualizations using R
- You don't have to learn complex tools, like D3
- You simply work in R learning a minimal amount of new syntaxt
- written by Ramnath Vaidyanathan (who also write Slidify)

```{r}
#require(devtools)
#install_github('rCharts', 'ramnathv')
library(rCharts)
workdir <- "C:\\Users\\clint\\GitHub\\LearnCode_R\\02_RPlot\\fig_rCharts"
```


```{r}
haireye = as.data.frame(HairEyeColor)
n1 <- nPlot(Freq ~ Hair, 
      group = 'Eye',
      type = 'multiBarChart',
      data = subset(haireye, Sex == 'Male'))
```

The object n1 contains the plot
- in RStudio, typing n1 brings up the plot in the RStudio viewer
- do n1$ then hit TAB to see the various functions contained in teh object
    - n1$html prints out the html for the plot
    - n1$save(filename) then bring the code back into slidify document
    
```{r}
r1 <- rPlot(Sepal.Length ~ Sepal.Width | Species,
            data = iris,
            color = 'Species',
            type = 'point')
r1
```


```{r}
r2 <-  rPlot(Freq ~ Hair | Eye,
             color = 'Eye', 
             data = haireye,
             type = 'bar')
r2
```


The next library we will be exploring is MorrisJS.

```{r}
data(economics, package = "ggplot2")
econ <- transform(economics, date = as.character(date))
m1 <- mPlot(x = "date", y = c("psavert", "uempmed"), type = "Line", data = econ)
m1$set(pointSize = 0, lineWidth = 1)
m1$print("chart2")
```

```{r}
m1
```

Leaflet
```{r}
map3 <- Leaflet$new()
map3$setView(c(51.505, -0.09), zoom = 13)
map3$marker(c(51.5, -0.09), bindPopup = "<p> Hi. I am a popup </p>")
map3$marker(c(51.495, -0.083), bindPopup = "<p> Hi. I am another popup </p>")
map3$print("chart7")
```

```{r}
map3
```

