---
title: "03 Function"
author: "Clint ko"
output: html_notebook
---

# Three Components of a function

- **body**        --- code
- **formals**     --- arguments
- **environment** --- where the function exists

Normally, a function contains these three components and it can be treated as an object.  
**Exception: Primitive function**

# Wickham's Principles

- **name masking**                     --- lexical scoping and environment
- **functions vs variables (no diff)** --- pass function as an object
- **fresh start**                      --- memoryless
- **dynamic lookups**                  --- how R search for the value of a variable

All function can lookup to the variable defined in global environment.

# Examples

**All function can lookup to the variable defined in global environment.**
```{r}
f <- function() x
x <- 20
cat("1st call:", f(), "\n")

x <- 10
cat("2nd call:", f(), "\n")
```

Test global and local variables
```{r}
f <- function(x){
    y <- x+2
}
f(2)        # assignment does not return things
print(f(2)) # it force the function to output the value
(f(2))      # it force the function to output the value
y <- 3      # assign globally
(f(2))      # name masking ---> inside f, there is a local variable 'y'
y           # globally
```

if there is a typo, you might get unexpected results
```{r}
Foobar <- 6
f <- function(foobar) {
    return (Foobar + 1)
}
f(4)
```

To avoid such errors, you can use findGlobals function in library codetools
```{r}
codetools::findGlobals(f)
```

**take away messageg: **  
make your function self-contain as possible, don't assume you will get values from global environment

- Note
    - functional programming
        - self-contained as possibile
        - no side effects (side effect: change values outside the environment of the function)
        - **some exception: par, setwd**
    - in compiling language, the compiler check the type of each variable
    - in scripting language, such check does not existc


# Calling functions
when calling functions, R looks for a function's:

- **named**
- **prefix**
- **positition**

```
# position variables
f1 <- function(a, b){
    a * (a + b)
}

f1(3, 4) # matched by position / order of arguments

# named variables
f2 <- function(a = 1, b = 2){
    a * (a + b)
}

f2(b = 2, a = 1) # matched by name

# prefix matching
f3 <- function(xx){
    return(xx)
}

f3(x = 3) # matched by prefix
```

- Note: 
    - use position when it is extremely obvious (ex: rnorm)
    - **dont ever use prefix matching**

```{r}
rnorm
# n        --- position
# mean, sd --- name
```

# Prefix & Infix

prefix
infix (ex: `+`)

```{r}
1 + 3
```

```{r}
`+`(1, 2)
```

To create your own infix function, use %<...>%