---
title: "04 Object"
author: "Clint ko"
output: html_notebook
---


<a name="Outline"></a>Outline

- [Programming Paradigms](#Paradigms)
    - [First/second generation](#1st2ndGeneration)
    - [Procedural Code](#ProceduralProgram)
    - [Object Programming](#ObjectProgram)
    - [Functional Programming](#FunctionalProgram)
    - [Literate Programming](#LiterateProgram)
- [Object in R](#ObjectR)
    - [S3 Object](#ObjectS3)
    - [S4 Object](#ObjectS4)
    - [Reference Class (RC) Object](#ObjectRC)
- [S3 Object](#S3)
- [S4 Object](#S4)
- [RC Object](#RC)



# <a name="Paradigms"></a>Programming Paradigms
[[Back to Outline]](#Outline)

Before talking about object programming in R, it is helpful to talk about types of programming language.

### <a name="1st2ndGeneration"></a>First/second generation
[[Back to Outline]](#Outline)

- Idea: 
    - you just want the machine to work
- Examples
    - machine code
    - assembly code 
        - [mnemonics](http://whatis.techtarget.com/definition/mnemonic)
        - labels for memory addresses
        - still used in embedded systems

### <a name="ProceduralProgram"></a>Procedural Programming
[[Back to Outline]](#Outline)

- Background
    - personal computers started to appear
- Idea
    - how code should be written => structured programming (loops, conditionals)
- Examples
    - **COBOL**
    - **Fortran**
    - **ALGOL**
        - Note: paragraph
    - **Beginners All purpose Symbolic Instruction Code (BASIC)**
        - Note: goto --- spagetti code
    - **C**
        - Note: structure

### <a name="ObjectProgram"></a>Object Programming
[[Back to Outline]](#Outline)

- Terms
    - encapsulation
    - classes
    - methohds
    - interaction
    - inheritance
    - objects
- Idea: class
    - contains both **data** and the **functions** that operate on it.
- Examples
    - **C++**
        - Note: struct (from C) + methods => Class
    
    
### <a name="FunctionalProgram"></a>Functional Programming
[[Back to Outline]](#Outline)

- Idea
    - keep functions pure as possible
    - keep self contained, etc
- Note
    - normally, you will always use something from global varaible
    ex: f <- function(a, b) a + b => '+' is a function from global environment

### <a name="LiterateProgram"></a>Literate Programming
[[Back to Outline]](#Outline)

- [Definitation](https://en.wikipedia.org/wiki/Literate_programming)

Literate programming is a programming paradigm introduced by Donald Knuth in which a program is given as an explanation of the program logic in a natural language, such as English, interspersed with snippets of macros and traditional source code, from which a compilable source code can be generated.

- Example
    - **Jupyter**
    - **R Notebook**

# <a name="ObjectR"></a>Object in R

To talk about object in R, it is important to know that there are three kinds of objects: S3, S4, and RC. They are created in different time in R history.

"S is R is an implementation of the S programming language. S is a statistical programming language developed primarily by John Chambers and (in earlier versions) Rick Becker and Allan Wilks of Bell Laboratories." (from wiki)

- *1976* S (**S**tatistical analysis software) 
    - Scheme / APL / PPL
        - [**Scheme**](https://en.wikipedia.org/wiki/Scheme_(programming_language)) --- functional programming
        - [**APL**](https://en.wikipedia.org/wiki/APL_(programming_language)) --- multi-dimensional array  
        - [**PPL**](https://en.wikipedia.org/wiki/Polymorphic_Programming_Language) --- interactive
    - Idea: Turn ideas into software quickly and faithfully
    - Version
        - OLD S: interactive way to call C + Fortran
        - New S: 
            - rewrite old Fortran using C 
            - concept of objects (**S3 objects**)
            - change macros into functions
        - S4  (last release of S)
            - provides "more advanced" object orientic features (**S4 objects**)
- *1993* R
    - base R is mostly S

### <a name="ObjectS3"></a>S3 Object
[[Back to Outline]](#Outline)

- just wrapping things together
- very flexible

### <a name="ObjectS4"></a>S4 Object
[[Back to Outline]](#Outline)

- more structured
- error checking

### <a name="ObjectRC"></a>Reference Class (RC) Object
[[Back to Outline]](#Outline)

- much like the object you have in other languages

# <a name="S3"></a>S3 Object
[[Back to Outline]](#Outline)

### Example: Summary function in data.frame & lm objects
```{r}
data(iris)
class(iris)
```

when applying to the summary of iris, it is actually calling "summary.data.frame" function
```{r}
summary(iris)
```

Here we can show that summary(iris) == summary.data.frame(iris)
```{r}
summary.data.frame(iris)
```

same thing go with lm object
```{r}
lm_iris <- lm(Sepal.Length ~ Sepal.Width, data = iris)
class(lm_iris)
```

Applying summary(lm object) == summary.lm(lm object)
```{r}
summary(lm_iris)
```

```{r}
summary.lm(lm_iris)
```

you can create your own summary function for a certain class; just create sumary.<name_of_class>
```{r}
summary.lm <- function(x){ print("haha") }
summary(lm_iris)
```


### Create S3 Objects
[[Back to Outline]](#Outline)

clean the environment
```{r}
rm(list = ls())
```

```{r}
Person <- function (x, y) {
  structure(list(name = x, age = y), class = "Person")
}

Employee <- function(x, y){
  structure(list(x, salary = y), class = "Employee")
}
```

to create a constructor , you use the function called "structure"
```{r}
Person <- function(x, y) {
    structure(
        list(name = x, age = y),
        class = "Person"
    ) # end struct
} # end func Person

Employee <- function(x, y) { # the input x will be a Person object
    structure(
        list(x, salary = y),
        class = "Employee"
    ) # end struct
} # end func Employee
```

create instance
```{r}
laura <- Person("Laura", 34)
laura_emp <- Employee(laura, 45)
```

output a Person object
```{r}
print(laura)
```

output a Employee object
```{r}
print(laura_emp)
```

create a generic function "Raise". What Raise will do is to call the Raise.<class_name>. (Note: what is [generic function](http://www.idc-online.com/technical_references/pdfs/information_technology/Generic_Functions_in_Python.pdf)?)
```{r}
Raise <- function(x) UseMethod("Raise", x)
#?UseMethod
```

<generic_function>.<class_name>
```{r}
Raise.Employee <- function(x){
   return(x$salary*(1.10))
}
```

```{r}
Raise.Employee(laura_emp)
```

```{r}
Raise(laura_emp)
```


# <a name="S4"></a>S4 Object
[[Back to Outline]](#Outline)

clean the environment
```{r}
rm(list = ls())
```

unlike S3, you do not need to create your own constructor. The function "setClass" will help you to create it
```{r}
setClass(
    "Person", 
    slots = list(
        name = "character", 
        age = "numeric"))

setClass(
    "Employee",
    slots = list(
        salary = "numeric"
    ), contains = "Person")
```

use function new to create instances
```{r}
alice <- new("Person", name = "Alice", age = 40)
john  <- new("Employee", name = "John", age = 20, salary = 40)
```

```{r}
alice
```

```{r}
john
```

To set generic function in S4 objects
```{r}
setGeneric(
    "Raise", 
    function(x) standardGeneric("Raise"))

setMethod(
    "Raise", 
    "Employee", 
    function(x) x@salary*(1.10))
```

```{r}
Raise(john)
```


# <a name="RC"></a>RC Object
[[Back to Outline]](#Outline)

```{r}
rm(list = ls())
```

In S3, the method is not really tide to the object. The only thing that connect them is the "name". On the other hand, RC object is a method passing object.

you use setRefClass to create a RC objects
```{r}
#setRefClass("Polygon", fields = c("sides"))
#setRefClass("Polygon", fields = list(sides = "numeric"))

Polygon <- setRefClass("Polygon", fields = c("sides"))
square <- Polygon$new(sides = 4)
triangle <- Polygon$new(sides = 4)
```


```{r}
print(Polygon)
```


```{r}
print(square)
```

```{r}
triangle$sides <- 3
print(triangle)
```

be careful about this
```{r}
triangle <- square
triangle$sides <- 3
```

```{r}
print(triangle)
```

however, not only the field in triangle is changed but also the field in square
```{r}
print(square)
```

To provide an example of passing reference
```{r}
x <- 4 # create x with a value 4
y <- x # pass the reference to y
```

```{r}
# the variable x and y has the same address
library(data.table)
print(address(x))
print(address(y))
```

- Note: 
    - the bioconductor mainly use S4 object

