---
title: "Foundation01_DataStructure"
output: html_document
---

# Intro to R data Structure

| Dimension | Homogenous    | Heterogeneous |
|-----------|---------------|---------------|
| 1d        | Atomic vector | List          |
| 2d        | Matrix        | Data frame    |
| nd        | Array         |               |


Note that R has no 0-dimensional, or scalar types. Individual numbers or strings, which you might think would be scalars, are actually vectors of length one.

Given an object, the best way to understand what data structures it? composed of is to use str(). **str() is short for structure** and it gives a compact, human readable description of any R data structure.

### Quiz
Take this short quiz to determine if you need to read this chapter. If the answers quickly come to mind, you can comfortably skip this chapter. You can check your answers in answers.


- *Q1* What are the three properties of a vector, other than its contents?
    - **MyAns:** value, names, length
    

- *Q2* What are the four common types of atomic vectors? What are the two rare types?
    - **MyAns:** 
        1. characterm, boolean, numeric, integer
        2. complex number, ???


- *Q3* What are attributes? How do you get them and set them?
    - **MyAns:** 
        - attributes are fields that belong to an object
        - attributes()


- *Q4* How is a list different from an atomic vector? How is a matrix different from a data frame?
    - **MyAns:** 
        - list is a collection of variables and each variables could be different types
        - matrix stores data of the same types, while columns in data frame could be composed of different types with each other


- *Q5* Can you have a list that is a matrix? Can a data frame have a column that is a matrix?
    - **MyAns:** 
        1. Yes
        2. No


### Outline
- Vectors introduces you to atomic vectors and lists, R?€™s 1d data structures.

- Attributes takes a small detour to discuss attributes, R?€™s flexible metadata specification. Here you?€™ll learn about factors, an important data structure created by setting attributes of an atomic vector.

- Matrices and arrays introduces matrices and arrays, data structures for storing 2d and higher dimensional data.

- Data frames teaches you about the data frame, the most important data structure for storing data in R. Data frames combine the behaviour of lists and matrices to make a structure ideally suited for the needs of statistical data.

# Vectors
- Vectors come in two flavours: atomic vectors and lists

- Three common properties
    - **Type**:       typeof(), what it is
    - **Length**:     length(), how many elements it contains
    - **Attributes**: attributes(), additional arbitrary metadata
    
- atomic vectors vs lists
    - atomic vectors: all elements must be the same type
    - list:           elements can have different types
    
## Vectors: Atomic vectors
- four common types
    - logical
    - integer
    - double (numeric)
    - character
    
- c(), short for combine
```{r}
dbl_var <- c(1, 2.5, 4.5)

# with the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)

# use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)

chr_var <- c("these are", "some strings")
```

Atomic vectors are always flat, even if you nest c()'s
```{r}
c(1, c(2, c(3, 4)))
#> [1] 1 2 3 4

# the same as
c(1, 2, 3, 4)
#> [1] 1 2 3 4
```

Missing value -> NA
- **NA_real_**
- **NA_integer_**
- **NA_character_**
```{r}
NA_real_
NA_integer_
NA_character_
```

### Types and tests
- typeof()
- "is" function
    - is.character()
    - is.double()
    - is.integer()
    - is.logical()
    - is.atomic()

```{r}
int_var <- c(1L, 6L, 10L)
typeof(int_var)
#> [1] "integer"
is.integer(int_var)
#> [1] TRUE
is.atomic(int_var)
#> [1] TRUE

dbl_var <- c(1.0, 2.5, 4.5)
typeof(dbl_var)
#> [1] "double"
is.double(dbl_var)
#> [1] TRUE
is.atomic(dbl_var)
#> [1] TRUE
```

is.numeric() is a general test for the "numberliness"
```{r}
is.numeric(int_var)
is.numeric(dbl_var)
```

### Coercion
different vector -> coerced to the most flexible type
least most flexible: logical, integer, double, and character
```{r}
str(c("a", 1))
#> chr [1:2] "a" "1"
```

when logical -> numeric
- TRUE  -> 1
- FALSE -> 0
useful in conjunction with sum() and mean()
```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
#> [1] 0 0 1

# Total number of TRUEs
sum(x)
#> [1] 1

# Proportion that are TRUE
mean(x)
#> [1] 0.333333
```

## Lists
- elements of list can be of any type, including list
- list()
```{r}
x <- list(1:3, "a", c(T, F, T), c(2.3, 5.9))
str(x)
#> List of 4
#>  $ : int [1:3] 1 2 3
#>  $ : chr "a"
#>  $ : logi [1:3] TRUE FALSE TRUE
#>  $ : num [1:2] 2.3 5.9
```

Lists are sometimes called recursive vectors, because a list can contain other lists. This makes them undamentally different from atomic vectors
```{r}
x <- list(list(list(list())))
str(x)
#> List of 1
#>   $ :List of 1
#>    ..$ :List of 1
#>    .. ..$ : list()

is.recursive(x)
#> [1] TRUE
```

c() will combine several lists into one. If given a combination of atomic vectors and lists, c() will coerce the vectors to lists before combining them. Compare the results of list() and c():
```{r}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
#> List of 2
#>  $ :List of 2
#>   ..$ : num 1
#>   ..$ : num 2
#>  $ : num [1:2] 3 4

str(y)
#> List of 4
#>  $ : num 1
#>  $ : num 2
#>  $ : num 3
#>  $ : num 4
```

typeof()
is.list()
as.list()
unlist()
```{r}
x <- c(1, 2, 3)
y <- list(1, 2, 3)
is.list(x)
is.list(y)
as.list(x)
unlist(y)
```

Lists are used to build up many of the more complicated data structures in R. For example, both data frames (described in data frames) and linear models objects (as produced by lm()) are lists:
```{r}
is.list(mtcars)
head(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

### Exercises
1. What are the six types of atomic vector? How does a list differ from an atomic vector?

2. What makes is.vector() and is.numeric() fundamentally differnt to is.list() and is.charactter()?

3. Test your knowledge of vector coercion rules by predicting the output of the following uses of c();
```{r}
c(1, FALSE)
c("a", 1)
c(list(1), "a") #!!!
c(T, 1L)
```

4. Why do you need to use unlist() to convert a list to an atomic vector? Why doesn't as.vector() work?
```{r}
x <- list(1, 2, 3)
as.vector(x)
as.integer(x)
c(x)
```

**MyAns**
because list is also a vector, therefore as.vector does not work

5. why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?

6. Why is the default missing value, NA, a logical vector? What's special about logical vectors? (Hint: think about c(FALSE, NA_character_)
```{r}
c(FALSE, NA_character_)
typeof(NA)
```

**MyAns**
because the logical is the least flexible types

## Attributes

- all objects can have arbitrary additional attributes
- store metadata about the object
- Attributes ---> named list (with unique names)
- attr()

```{r}
# example: atomic vector
y <- 1:10

# assigned attributes
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")

# show the attributes
y
attributes(y)
str(attributes(y))
```

The structure() function returns a new object with modified attributes:
```{r}
structure(1:10, my_attribute = "This is a vector")
```

By default, most attributes are lost when modifying a vector:
```{r}
attributes(y[1])
attributes(sum(y))
```

The only attributes not lost are three most important
- names
    - a character vector giving each element a name
    
- Dimensions
    - used to turn vectors into matrices and arrays
    
- class
    - used to implement the S3 object system
    
functions that get and set those attributes
- names()
- dim()
- class()
- not attr(x, "names"), attr(x, "dim"), and attr(x, "class")

### Names
name vector --- three ways
- when creating it
    - x <- c(a = 1, b = 2, c = 3)
    
- By modifying an existing vector in place
    - x <- 1:3
    - names(x) <- c("a", "b", "c")
    
- By creating a modified copy of a vector
    - x <- setNames(1:3, c("a", "b", "c"))

- Names do not have to be unique        
    - character subsetting si the most important reason to use names
    - it is most useful when the names are unique
    
```{r}
y <- c(a = 1, 2, 3)
names(y)


z <- c(1, 2, 3)
names(z)

x <- c(a = 1, b = 2, c = 3)
names(x)
x["a"]

x <- c(a = 1, a = 2, c = 3)
names(x)
x["a"]
```


You can create a new vector without names using unname(x), or remove names in place with names(x) <- NULL
```{r}
x <- c(a = 1, a = 2, c = 3)
unname(x)
x

x <- c(a = 1, a = 2, c = 3)
names(x) <- NULL
x
```


## Factors
a factor a vector that can contain only predefined values, and is used to store categorical data

Two attributes
- class()
- levels()

```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)
```

you cannot use values that are not in the levels
```{r}
x[2] <- "c"
x
```

you cannot combine factors
```{r}
c(factor("a"), factor("b"))
```

Factors are useful when you know the possible values a variable may take, even if you don¡¦t see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:
```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Sometime when a data frame is read directly from a file
-> columns are often factors instead of integers
-> this is often due na string in the data

this causes problem when reading values 
```{r}
z <- read.csv(text = "value\n12\n1\n.\n9")
typeof(z$value)
class(z$value)
as.double(z$value) # we got the levels of a factor
```

to solve the problem
```{r}
z <- read.csv(text = "value\n12\n1\n.\n9", na.strings=".")
typeof(z$value)
class(z$value)
z$value
```


Unfortunately, most data loading functions in R automatically convert character vectors to factors. This is suboptimal, because there¡¦s no way for those functions to know the set of all possible levels or their optimal order. Instead, use the argument stringsAsFactors = FALSE to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the data. A global option, options(stringsAsFactors = FALSE), is available to control this behaviour, but I don¡¦t recommend using it. Changing a global option may have unexpected consequences when combined with other code (either from packages, or code that you¡¦re source()ing), and global options make code harder to understand because they increase the number of lines you need to read to understand how a single line of code will behave.


**While factors look (and often behave) like character vectors, they are actually integers.** Be careful when treating them like strings. Some string methods (like gsub() and grepl()) will coerce factors to strings, while others (like nchar()) will throw an error, and still others (like c()) will use the underlying integer values. For this reason, it¡¦s usually best to explicitly convert factors to character vectors if you need string-like behaviour. In early versions of R, there was a memory advantage to using factors instead of character vectors, but this is no longer the case.
```{r}
x <- factor(letters)
str(x)
unclass(x)
```


### Exercise
**Q1**
An early draft used this code to illustrate structure():
```{r}
structure(1:10, my_attribute = "This is a vector")
```

however, run the follow code
```{r}
structure(1:5, comment = "my attribute")
```

Why? Is the attribute missing
```{r}
structure(1:5, comment = "my attribute")
structure(1:5, comments = "my attribute")
structure(1:5, list = "my attribute")
```

```{r}
help(structure)
help(comment)
```

```{r}
x <- structure(1:5, comment = "my attribute")
comment(x)
```

**Q2**
What happens to a factor when you modify its levels?
```{r}
f1 <- factor(letters)
f1

levels(f1) <- rev(levels(f1))
f1
```

**Q3**
What does this code do? How do f2 and f3 differ from f1?
```{r}
f1 <- factor(letters)
f1

f2 <- rev(factor(letters))
f2

f3 <- factor(letters, levels = rev(letters))
f3
```

MyAns:
```{r}
labels(f1)
labels(f2)
labels(f3)

levels(f1)
levels(f2)
levels(f3)
```

## Matrices and arrays